---
title: "Penguin Body Size in R"
author: "Indervir Dhillon"
date: 2026-09-23
---

How do average body mass and flipper length differ by penguin species
and sex?

Data: [Palmer Penguins](https://allisonhorst.github.io/palmerpenguins/),
collected by Kristen Gorman and Palmer Station Antarctica LTER.
The data are available under the
[CC0 licence](https://allisonhorst.github.io/palmerpenguins/LICENSE.html).

Citation: Horst, A. M., Hill, A. P., & Gorman, K. B. (2020).
*palmerpenguins: Palmer Archipelago (Antarctica) penguin data*.
[https://doi.org/10.5281/zenodo.3960218](https://doi.org/10.5281/zenodo.3960218).

```{r}
#| label: setup
#| message: false
#| warning: false

library(palmerpenguins)
library(dplyr)
library(ggplot2)
```

## Check the data

```{r}
penguins <- palmerpenguins::penguins
columns <- c("species", "sex", "body_mass_g", "flipper_length_mm")

print(paste("Number of penguins:", nrow(penguins)))
print("Missing values in the columns I will use:")
print(colSums(is.na(penguins[columns])))
```

Some rows are missing a measurement or a sex value. I check these
before comparing groups.

## Remove incomplete rows

```{r}
clean <- penguins[complete.cases(penguins[columns]), ]

print(paste("Rows kept:", nrow(clean)))
print(paste("Rows removed:", nrow(penguins) - nrow(clean)))
print(head(clean[columns]))
```

I use the same complete rows for the summary and plot.

## Calculate group averages

```{r}
summary <- clean |>
  group_by(species, sex) |>
  summarise(
    count = n(),
    mean_mass_g = mean(body_mass_g),
    mean_flipper_mm = mean(flipper_length_mm),
    .groups = "drop"
  )

print(as.data.frame(summary), row.names = FALSE)
```

Gentoo males have the highest average body mass of 5484.8 g, while
Adelie females have the lowest average body mass of 3368.8 g. Within
each species, males have a higher average body mass than females.

Gentoo males have the longest average flipper length of 221.5 mm, 
while Adelie females have the shortest of 187.8 mm.
Within each species, similarly, males have longer average flipper
lengths than females.


## Plot average body mass

```{r}
#| label: fig-r-penguin-average-mass
#| fig-cap: "Average body mass by penguin species and sex."

ggplot(summary, aes(x = species, y = mean_mass_g, fill = sex)) +
  geom_col(position = "dodge") +
  labs(
    x = "Species",
    y = "Average body mass (g)",
    fill = "Sex"
  ) +
  theme_minimal()
```

The plot here shows the average body mass by penguin species and sex.
We can see that Gentoo penguins have the highest average body mass in
this sample.

